---
title: "Breast Cancer Wisconsin Diagnostic — Exploratory Analysis"
author: "Somnath Chaudhuri"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  pdf_document:
    latex_engine: xelatex
    toc: true
    number_sections: true
  html_document:
    toc: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 7, fig.height = 5)

# Load required libraries
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(GGally)
library(DataExplorer)
library(patchwork)
library(corrplot)
library(factoextra)
library(reshape2)
```

# Introduction

This short training document shows step-by-step exploratory data analysis (EDA) for the **Breast Cancer Wisconsin (Diagnostic)** dataset.  
The goal is to demonstrate a compact, reproducible workflow with clear interpretations before modeling.

# Data loading and structure

```{r data-load}
# Load data and column names
breastCancerData <- read_csv(
  "https://archive.ics.uci.edu/ml/machine-learning-databases/breast-cancer-wisconsin/wdbc.data",
  col_names = FALSE
)

breastCancerDataColNames <- read_csv(
  "https://raw.githubusercontent.com/fpsom/2020-07-machine-learning-sib/master/data/wdbc.colnames.csv",
  col_names = FALSE
)

# Check counts match
ncol(breastCancerData)
nrow(breastCancerDataColNames)

# Assign names safely
if (nrow(breastCancerDataColNames) == ncol(breastCancerData)) {
  colnames(breastCancerData) <- breastCancerDataColNames$X1
} else {
  stop("Column name count does not match data columns. Please verify column source.")
}

# Clean column names for safe reference
colnames(breastCancerData) <- make.names(colnames(breastCancerData))

# Convert diagnosis to factor
breastCancerData$Diagnosis <- as.factor(breastCancerData$Diagnosis)

# Remove first column
breastCancerData <- breastCancerData[2:ncol(breastCancerData)]

# Quick checks
dim(breastCancerData)
names(breastCancerData)
head(breastCancerData)
summary(breastCancerData$diagnosis)
```

**Inference:** The dataset has 569 observations and 32 variables. The response variable `diagnosis` has two levels: *M (Malignant)* and *B (Benign)*.

# Missingness and overview

```{r data-intro}
plot_intro(breastCancerData)
plot_missing(breastCancerData)
```

**Observation:** No missing data — the dataset is complete and balanced for analysis.


# Using first 6 columns of the dataset to explore

```{r drop}
breastCancerData <- breastCancerData%>% select(1:6)
# Quick checks
dim(breastCancerData)
names(breastCancerData)
head(breastCancerData)
```

# Univariate distributions

```{r univariate}
num_cols <- breastCancerData %>% select(where(is.numeric))
long_df <- num_cols %>%
pivot_longer(cols = everything(), names_to = "variable", values_to = "value")

p_density <- ggplot(long_df, aes(x = value, fill = variable)) +
geom_density(alpha = 0.5) +
facet_wrap(~variable, scales = "free", ncol = 3) +
labs(title = "Density plots for selected numeric variables") +
theme_minimal() +
theme(legend.position = "none")

p_box <- ggplot(long_df, aes(x = variable, y = value, fill = variable)) +
geom_boxplot(alpha = 0.7) +
labs(title = "Boxplots (overview)", x = NULL) +
theme_minimal() +
theme(axis.text.x = element_blank(), legend.position = "none")

p_density / p_box

```

**Inference:** Some variables (e.g., radius_mean and area_mean) are right-skewed,
suggesting the need for scaling or transformation before modeling.

# Bivariate relationships by diagnosis

```{r bivariate}
p1 <- ggplot(breastCancerData, aes(x = Radius.Mean, y = Area.Mean, color = Diagnosis)) +
geom_point(size = 2.5, alpha = 0.8) +
geom_smooth(method = "lm", se = FALSE, size = 0.6) +
labs(title = "Radius vs Area by Diagnosis") +
theme_minimal()

p2 <- ggplot(breastCancerData, aes(x = Texture.Mean, y = Smoothness.Mean, color = Diagnosis)) +
geom_point(size = 2.5, alpha = 0.8) +
geom_smooth(method = "lm", se = FALSE, size = 0.6) +
labs(title = "Texture vs Smoothness by Diagnosis") +
theme_minimal()

p1 / p2

```

**Inference:** Malignant tumors tend to have higher radius, area, and texture values.
Benign tumors generally have lower measurements and smoother surfaces.

# Distribution by diagnosis (violin plots)

```{r violin}
df_long <- breastCancerData %>%
pivot_longer(-Diagnosis, names_to = "variable", values_to = "value")

ggplot(df_long, aes(x = Diagnosis, y = value, fill = Diagnosis)) +
geom_violin(alpha = 0.5, trim = FALSE) +
geom_boxplot(width = 0.15, outlier.size = 1, alpha = 0.9) +
facet_wrap(~variable, scales = "free", ncol = 3) +
labs(title = "Variable distribution by Diagnosis", x = "Diagnosis", y = "Value") +
theme_minimal() +
theme(legend.position = "none")
```

**Inference:** Malignant cases show higher radius, perimeter, area, and texture. Smoothness is slightly higher for malignant tumors.

# Correlation matrix

```{r corr}
num <- breastCancerData %>% select(where(is.numeric))
M <- cor(num)
corrplot(M, method = "pie")
round(M, 2)
```

**Inference:** Radius, perimeter, and area are strongly correlated. Texture and smoothness are moderately correlated with these measures.

# Pairwise scatter matrix

```{r ggpairs}
ggpairs(
breastCancerData,
columns = 2:6,
mapping = aes(color = Diagnosis, alpha = 0.6),
upper = list(continuous = wrap("cor", size = 3)),
lower = list(continuous = wrap("smooth", alpha = 0.3, size = 0.4)),
diag = list(continuous = wrap("densityDiag", alpha = 0.5))
)
```

**Inference:** Malignant tumors cluster in the higher ranges for radius, perimeter, and area.
Benign tumors occupy lower ranges. Strong correlations exist among size-related features.

# Feature engineering & scaling

```{r features}
# Example: area-to-radius ratio
breastCancerData <- breastCancerData %>%
mutate(area_to_radius = Area.Mean / Radius.Mean)

summary(breastCancerData$area_to_radius)

# Scale numeric features
scaled_data <- as.data.frame(scale(breastCancerData %>% select(where(is.numeric))))
scaled_data$Diagnosis <- breastCancerData$Diagnosis

head(scaled_data)
```

**Inference:** Derived ratio captures proportionality differences.
Scaled data ensures features contribute equally for modeling.

# Pre-modelling checklist

1. Classes moderately balanced between Malignant and Benign.
2. No missing values.  
3. High correlations among radius, area, and perimeter — consider dimensionality reduction.  
4. Scaling and derived ratios improve comparability across features.  

