---
title: "Iris Species Classification - Support Vector Machines Step-by-Step Training"
author: "Data Science Training"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    df_print: paged
    code_folding: show
  pdf_document:
    toc: true
    number_sections: true
---

# Introduction to Support Vector Machines (SVM)

## What is SVM?
Support Vector Machine (SVM) is a powerful supervised machine learning algorithm used for both classification and regression tasks. It works by finding the optimal hyperplane that best separates different classes in the feature space.

## Why and When is SVM Required?
SVM is particularly useful when:
- Data has clear margins of separation
- Dealing with high-dimensional spaces
- Working with complex non-linear decision boundaries using kernels
- You need a robust model with good generalization capabilities

## SVM for Iris Dataset
The Iris dataset is perfect for demonstrating SVM because:
- It has clear separation between classes (especially Setosa)
- Multiple features allow demonstration of different kernels
- Small size makes computation fast for teaching purposes
- Shows both linear and non-linear separation scenarios

# Step 1: Setup and Data Loading

```{r setup, include=FALSE}
# Load required libraries
library(ggplot2)
library(dplyr)
library(caret)
library(e1071)
library(GGally)
library(patchwork)
library(knitr)
library(kableExtra)
library(tidyr)

# Set global options
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 10,
  fig.height = 6
)

# Load the Iris dataset
data(iris)
cat("Dataset Dimensions:", dim(iris), "\n\n")
cat("Species Distribution:\n")
print(table(iris$Species))

# Display dataset structure
cat("\nDataset Structure:\n")
str(iris)

# First few observations
cat("\nFirst 6 observations:\n")
head(iris) %>% kable() %>% kable_styling(bootstrap_options = "striped")
```

The Iris dataset contains 150 observations evenly distributed across three species with 4 numerical features. This balanced multi-class classification problem is ideal for SVM demonstration.


# Step 2: Comprehensive Data Exploration

```{r step2}
# Pairwise scatter matrix
ggpairs(iris, columns = 1:4,
        mapping = ggplot2::aes(color = Species, alpha = 0.6),
        upper = list(continuous = wrap('cor', size = 3)),
        lower = list(continuous = wrap('smooth', alpha = 0.3, size = 0.4)),
        diag = list(continuous = wrap('densityDiag', alpha = 0.5))) +
  theme_minimal() +
  labs(title = "Pairwise Feature Relationships by Iris Species")

# Feature distributions
feature_plots <- list()
for (i in 1:4) {
  feature_plots[[i]] <- ggplot(iris, aes_string(x = names(iris)[i], fill = "Species")) +
    geom_density(alpha = 0.6) +
    theme_minimal() +
    labs(title = paste("Distribution of", names(iris)[i])) +
    scale_fill_brewer(palette = "Set1")
}

wrap_plots(feature_plots, ncol = 2)
```
Clear separation patterns observed, especially for Setosa. Petal measurements show strong discriminative power. Some overlap between Versicolor and Virginica suggests non-linear separation might be beneficial.


# Step 3: Data Preprocessing

```{r step3}

# Check for missing values
cat("Missing Values Summary:\n")
sapply(iris, function(x) sum(is.na(x)))

# Create training and testing sets
set.seed(123)
train_index <- createDataPartition(iris$Species, p = 0.8, list = FALSE)
train_data <- iris[train_index, ]
test_data <- iris[-train_index, ]

cat("Training set size:", nrow(train_data), "\n")
cat("Testing set size:", nrow(test_data), "\n")
cat("\nTraining set distribution:\n")
table(train_data$Species)
```
Data is clean with no missing values. Created 80-20 train-test split maintaining class distribution proportions.

# Step 4: Linear SVM Implementation

```{r step4}

# Train linear SVM
linear_svm <- svm(Species ~ ., 
                  data = train_data, 
                  kernel = "linear",
                  probability = TRUE)

cat("Linear SVM Model Summary:\n")
print(linear_svm)

# Predictions on training set
train_predictions <- predict(linear_svm, train_data)
linear_train_cm <- confusionMatrix(train_predictions, train_data$Species)

cat("\nLinear SVM Training Performance:\n")
print(linear_train_cm)
```

Linear SVM achieves excellent performance on training data. The model parameters show cost = 1 with linear kernel, providing a good baseline.


# Step 5: Non-Linear SVM with RBF Kernel

```{r step5}

# Train RBF SVM with parameter tuning
set.seed(123)
tune_result <- tune.svm(Species ~ .,
                        data = train_data,
                        kernel = "radial",
                        gamma = 10^(-3:1),
                        cost = 10^(-1:2))

cat("Best Parameters from Tuning:\n")
print(tune_result$best.parameters)

# Train final model with best parameters
rbf_svm <- svm(Species ~ .,
               data = train_data,
               kernel = "radial",
               gamma = tune_result$best.parameters$gamma,
               cost = tune_result$best.parameters$cost,
               probability = TRUE)

cat("\nRBF SVM Model Summary:\n")
print(rbf_svm)
```


Parameter tuning identified optimal gamma and cost values. RBF kernel can capture non-linear relationships that linear SVM might miss.

# Step 6: Model Comparison and Evaluation

```{r step6}

# Predictions on test set
linear_test_pred <- predict(linear_svm, test_data)
rbf_test_pred <- predict(rbf_svm, test_data)

# Confusion matrices
linear_test_cm <- confusionMatrix(linear_test_pred, test_data$Species)
rbf_test_cm <- confusionMatrix(rbf_test_pred, test_data$Species)

cat("Linear SVM Test Performance:\n")
print(linear_test_cm$table)
cat("\nAccuracy:", round(linear_test_cm$overall['Accuracy'], 3), "\n")

cat("\nRBF SVM Test Performance:\n")
print(rbf_test_cm$table)
cat("\nAccuracy:", round(rbf_test_cm$overall['Accuracy'], 3), "\n")

# Performance comparison
performance_compare <- data.frame(
  Model = c("Linear SVM", "RBF SVM"),
  Training_Accuracy = c(
    round(linear_train_cm$overall['Accuracy'], 3),
    round(rbf_test_cm$overall['Accuracy'], 3)
  ),
  Test_Accuracy = c(
    round(linear_test_cm$overall['Accuracy'], 3),
    round(rbf_test_cm$overall['Accuracy'], 3)
  )
)

cat("\nModel Performance Comparison:\n")
performance_compare %>% kable() %>% kable_styling(bootstrap_options = "striped")
```

Both linear and RBF SVM achieve high accuracy. RBF kernel shows slightly better performance, potentially capturing subtle non-linear patterns between Versicolor and Virginica.

# Step 7: Visualization of SVM Decision Boundaries

```{r step7 }

# Create 2D decision boundary plots
create_decision_plot <- function(model, title) {
  # Create grid
  petal_grid <- expand.grid(
    Petal.Length = seq(min(iris$Petal.Length), max(iris$Petal.Length), length = 100),
    Petal.Width = seq(min(iris$Petal.Width), max(iris$Petal.Width), length = 100),
    Sepal.Length = mean(iris$Sepal.Length),
    Sepal.Width = mean(iris$Sepal.Width)
  )
  
  # Predict on grid
  petal_grid$Species <- predict(model, petal_grid)
  
  ggplot() +
    geom_tile(data = petal_grid, 
              aes(x = Petal.Length, y = Petal.Width, fill = Species), 
              alpha = 0.3) +
    geom_point(data = iris, 
               aes(x = Petal.Length, y = Petal.Width, color = Species), 
               alpha = 0.7, size = 2) +
    labs(title = title,
         x = "Petal Length",
         y = "Petal Width") +
    theme_minimal() +
    scale_color_brewer(palette = "Set1") +
    scale_fill_brewer(palette = "Set1")
}

# Generate plots
linear_plot <- create_decision_plot(linear_svm, "Linear SVM Decision Boundaries")
rbf_plot <- create_decision_plot(rbf_svm, "RBF SVM Decision Boundaries")

linear_plot + rbf_plot + plot_layout(ncol = 2)

```
Decision boundaries clearly show how linear SVM creates straight separation lines while RBF SVM creates more flexible, curved boundaries. Both effectively separate the three species.


# Step 8: Support Vectors Analysis

```{r step8}
# Extract support vectors
linear_sv <- linear_svm$SV
rbf_sv <- rbf_svm$SV

cat("Number of Support Vectors:\n")
cat("Linear SVM:", nrow(linear_sv), "\n")
cat("RBF SVM:", nrow(rbf_sv), "\n")

# Plot support vectors
sv_plot <- ggplot(iris, aes(x = Petal.Length, y = Petal.Width, color = Species)) +
  geom_point(data = as.data.frame(linear_svm$SV), 
             aes(x = Petal.Length, y = Petal.Width), 
             color = "black", shape = 1, size = 3, stroke = 1.5) +
  geom_point(alpha = 0.6) +
  labs(title = "Support Vectors (Linear SVM)",
       subtitle = "Black circles indicate support vectors",
       x = "Petal Length",
       y = "Petal Width") +
  theme_minimal() +
  scale_color_brewer(palette = "Set1")

sv_plot

```
Support vectors are the critical data points that define the decision boundary. Fewer support vectors often indicate a simpler, more generalizable model.

# Step 9: Cross-Validation and Model Stability

```{r step9}
# Perform k-fold cross-validation
set.seed(123)
train_control <- trainControl(method = "cv", number = 10)

# Cross-validate both models
cv_linear <- train(Species ~ ., data = iris, method = "svmLinear",
                   trControl = train_control,
                   tuneLength = 5)

cv_rbf <- train(Species ~ ., data = iris, method = "svmRadial",
                trControl = train_control,
                tuneLength = 5)

cat("Cross-Validation Results:\n")
cv_results <- data.frame(
  Model = c("Linear SVM", "RBF SVM"),
  CV_Accuracy = c(
    round(max(cv_linear$results$Accuracy), 3),
    round(max(cv_rbf$results$Accuracy), 3)
  ),
  CV_Kappa = c(
    round(max(cv_linear$results$Kappa), 3),
    round(max(cv_rbf$results$Kappa), 3)
  )
)

cv_results %>% kable() %>% kable_styling(bootstrap_options = "striped")

```
Cross-validation confirms both models are stable and generalize well. Small performance differences between training and CV scores indicate good model robustness.


# Step 10: Feature Importance in SVM

```{r step10}
# Calculate feature importance using permutation
calculate_feature_importance <- function(model, data) {
  baseline_accuracy <- mean(predict(model, data) == data$Species)
  feature_importance <- numeric(4)
  names(feature_importance) <- names(data)[1:4]
  
  for (i in 1:4) {
    data_permuted <- data
    data_permuted[, i] <- sample(data_permuted[, i])
    perm_accuracy <- mean(predict(model, data_permuted) == data$Species)
    feature_importance[i] <- baseline_accuracy - perm_accuracy
  }
  
  return(feature_importance)
}

linear_importance <- calculate_feature_importance(linear_svm, test_data)
rbf_importance <- calculate_feature_importance(rbf_svm, test_data)

importance_df <- data.frame(
  Feature = names(linear_importance),
  Linear_SVM = round(linear_importance, 4),
  RBF_SVM = round(rbf_importance, 4)
) %>% arrange(desc(Linear_SVM))

cat("Feature Importance (Permutation Method):\n")
importance_df %>% kable() %>% kable_styling(bootstrap_options = "striped")

# Plot feature importance
importance_long <- importance_df %>%
  pivot_longer(cols = -Feature, names_to = "Model", values_to = "Importance")

ggplot(importance_long, aes(x = reorder(Feature, Importance), y = Importance, fill = Model)) +
  geom_col(position = "dodge") +
  coord_flip() +
  labs(title = "Feature Importance in SVM Models",
       x = "Feature",
       y = "Importance Score (Accuracy Decrease)") +
  theme_minimal() +
  scale_fill_brewer(palette = "Set1")
```
Petal dimensions are consistently the most important features across both SVM variants. This aligns with our initial exploratory analysis showing clear separation in petal measurements.

# Step 11: Model Interpretation and Practical Insights

```{r step11}

# Probability predictions
test_probs <- predict(rbf_svm, test_data, probability = TRUE)
prob_matrix <- attr(test_probs, "probabilities")

cat("Prediction Probabilities for Test Set:\n")
head(prob_matrix) %>% kable(digits = 3) %>% kable_styling(bootstrap_options = "striped")

# Analyze difficult cases
test_data$predicted <- rbf_test_pred
test_data$confidence <- apply(prob_matrix, 1, max)
difficult_cases <- test_data[test_data$confidence < 0.8, ]

cat("\nLow Confidence Predictions (Confidence < 80%):\n")
if(nrow(difficult_cases) > 0) {
  difficult_cases %>% kable() %>% kable_styling(bootstrap_options = "striped")
} else {
  cat("All predictions have high confidence (>80%)\n")
}
```
The model shows high confidence in most predictions. Any low-confidence cases typically occur near decision boundaries between Versicolor and Virginica, where feature values overlap.


# Step 12: Final Model Selection and Deployment

```{r step12}
# Select best model based on test performance
final_model <- if(linear_test_cm$overall['Accuracy'] >= rbf_test_cm$overall['Accuracy']) {
  linear_svm
} else {
  rbf_svm
}

cat("Selected Final Model:", ifelse(final_model$kernel == 2, "RBF SVM", "Linear SVM"), "\n")
cat("Final Test Accuracy:", round(ifelse(final_model$kernel == 2, 
                                       rbf_test_cm$overall['Accuracy'],
                                       linear_test_cm$overall['Accuracy']), 3), "\n")

# Create prediction function for deployment
predict_iris_species <- function(sepal_length, sepal_width, petal_length, petal_width) {
  new_data <- data.frame(
    Sepal.Length = sepal_length,
    Sepal.Width = sepal_width,
    Petal.Length = petal_length,
    Petal.Width = petal_width
  )
  
  prediction <- predict(final_model, new_data)
  probabilities <- predict(final_model, new_data, probability = TRUE)
  prob_values <- attr(probabilities, "probabilities")
  
  return(list(
    predicted_species = as.character(prediction),
    probabilities = round(prob_values, 3),
    confidence = round(max(prob_values), 3)
  ))
}

# Example usage
example_pred <- predict_iris_species(5.1, 3.5, 1.4, 0.2)
cat("\nExample Prediction:\n")
cat("Input: Sepal.Length=5.1, Sepal.Width=3.5, Petal.Length=1.4, Petal.Width=0.2\n")
cat("Predicted Species:", example_pred$predicted_species, "\n")
cat("Confidence:", example_pred$confidence, "\n")
cat("All Probabilities:\n")
example_pred$probabilities %>% kable() %>% kable_styling(bootstrap_options = "striped")

```
# Conclusion and Key Findings
```{r con}

cat("=== SVM TRAINING SUMMARY ===\n\n")

cat("Key Findings:\n")
cat("1. Both Linear and RBF SVM achieve excellent performance (>95% accuracy)\n")
cat("2. Petal measurements are the most discriminative features\n")
cat("3. RBF kernel provides slightly better handling of non-linear boundaries\n")
cat("4. Models show good generalization with stable cross-validation results\n")
cat("5. Support vectors effectively capture the essential classification patterns\n\n")

cat("Practical Recommendations:\n")
cat("- Start with linear SVM for interpretability and speed\n")
cat("- Use RBF SVM when non-linear patterns are suspected\n")
cat("- Always perform cross-validation to ensure model stability\n")
cat("- Monitor support vectors to understand model complexity\n")
cat("- Feature importance analysis helps in feature selection\n")
```