---
title: "Decision Tree vs Random Forest: Comprehensive Comparison"
author: "Data Science Training"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    df_print: paged
    code_folding: show
  pdf_document:
    toc: true
    number_sections: true
---



# Introduction to Decision Trees and Random Forests
## What are Decision Trees?
Decision Trees are supervised learning algorithms that make predictions by learning simple decision rules inferred from data features. Think of it as playing "20 Questions" — each node in the tree asks a question about a feature, and based on the answer, you follow the appropriate branch until you reach a prediction.

**Real-world example: A bank deciding whether to approve a loan:**

- Is income > $50,000?  
  - Yes: Is credit score > 700?  
    - Yes: Approve loan  
    - No: Reject loan  
  - No: Reject loan  

## What are Random Forests?
Random Forest is an ensemble method that combines multiple decision trees to create a more robust and accurate model. It uses two key techniques: bagging (bootstrap aggregating) and feature randomness. Instead of relying on a single tree, it builds many trees and combines their predictions.

**Real-world example: Medical diagnosis by a team of doctors:**

- Each specialist (tree) examines different aspects (features)  
- They may look at different patient data (feature randomness)  
- Each gives their diagnosis  
- Final diagnosis is based on majority vote 

Random Forests are typically more accurate and robust than a single Decision Tree but at the cost of reduced interpretability and higher computational demand.

## Why Compare Both Methods on the Same Dataset?
This comparison demonstrates key machine learning concepts:

- **Bias-Variance Tradeoff**: Single trees have high variance, forests reduce it  
- **Ensemble Learning**: Combining weak learners creates strong learners  
- **Overfitting**: Trees often overfit, forests generalize better  
- **Feature Importance**: Both methods provide interpretable feature rankings  
- **Computational Complexity**: Trade-off between simplicity and performance  

The choice between them depends on project requirements: use Decision Trees when explainability is crucial and relationships are simple, and Random Forest when maximum accuracy and robustness are needed for complex patterns. Both methods provide valuable feature importance insights that can inform domain understanding and feature engineering.


# Step 1: Setup and Data Loading
```{r}
# Load required libraries
library(ggplot2)
library(dplyr)
library(caret)
library(randomForest)
library(rpart)
library(rpart.plot)
library(GGally)
library(patchwork)
library(knitr)
library(kableExtra)
library(tidyr)
library(pROC)

# Load the Heart Disease dataset
url <- "https://raw.githubusercontent.com/sharmaroshan/Heart-UCI-Dataset/master/heart.csv"
heart <- read.csv(url)

cat("Dataset Dimensions:", dim(heart), "\n\n")
cat("Target Variable Distribution:\n")
print(table(heart$target))

# Display dataset structure
cat("\nDataset Structure:\n")
str(heart)

# First few observations
cat("\nFirst 6 observations:\n")
head(heart) %>% kable() %>% kable_styling(bootstrap_options = "striped")
```

Purpose: This block loads the heart disease dataset and provides initial exploration. We're using a real medical dataset to make our comparison more meaningful and practical.


Inference: The dataset contains 303 observations with 14 features including clinical parameters. The target variable is balanced (165 with heart disease, 138 without), making it suitable for binary classification without needing extensive class balancing.



# Step 2: Comprehensive Data Exploration
```{r exp, warning=FALSE}
# Basic statistics summary
cat("Summary Statistics:\n")
summary(heart) %>% kable() %>% kable_styling(bootstrap_options = "striped")

# Correlation heatmap
numeric_vars <- heart %>% select(where(is.numeric))
cor_matrix <- cor(numeric_vars)

cor_df <- as.data.frame(as.table(cor_matrix))
names(cor_df) <- c("Var1", "Var2", "Correlation")

ggplot(cor_df, aes(Var1, Var2, fill = Correlation)) +
  geom_tile() +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", 
                       midpoint = 0, limit = c(-1,1), space = "Lab", 
                       name="Correlation") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Feature Correlation Heatmap",
       x = "", y = "") +
  geom_text(aes(label = round(Correlation, 2)), color = "black", size = 3)

# Target distribution by key features
p1 <- ggplot(heart, aes(x = factor(target), y = age, fill = factor(target))) +
  geom_boxplot(alpha = 0.7) +
  labs(title = "Age Distribution by Heart Disease Status",
       x = "Heart Disease (0=No, 1=Yes)", y = "Age") +
  theme_minimal()

p2 <- ggplot(heart, aes(x = factor(target), y = chol, fill = factor(target))) +
  geom_boxplot(alpha = 0.7) +
  labs(title = "Cholesterol Distribution by Heart Disease Status",
       x = "Heart Disease (0=No, 1=Yes)", y = "Cholesterol") +
  theme_minimal()

p3 <- ggplot(heart, aes(x = factor(sex), fill = factor(target))) +
  geom_bar(position = "fill") +
  labs(title = "Heart Disease Prevalence by Gender",
       x = "Sex (0=Female, 1=Male)", y = "Proportion") +
  scale_fill_discrete(name = "Heart Disease") +
  theme_minimal()

p4 <- ggplot(heart, aes(x = factor(cp), fill = factor(target))) +
  geom_bar(position = "fill") +
  labs(title = "Heart Disease by Chest Pain Type",
       x = "Chest Pain Type", y = "Proportion") +
  scale_fill_discrete(name = "Heart Disease") +
  theme_minimal()

(p1 + p2) / (p3 + p4)

# Pairwise scatter matrix for Heart Disease dataset
ggpairs(heart, columns = c("age", "trestbps", "chol", "thalach", "oldpeak"),
        mapping = ggplot2::aes(color = factor(target), alpha = 0.6),
        upper = list(continuous = wrap('cor', size = 3)),
        lower = list(continuous = wrap('smooth', alpha = 0.3, size = 0.4)),
        diag = list(continuous = wrap('densityDiag', alpha = 0.5))) +
  theme_minimal() +
  labs(title = "Pairwise Feature Relationships by Heart Disease Status",
       subtitle = "Target: 0 = No Heart Disease, 1 = Heart Disease") +
  scale_color_manual(name = "Heart Disease", 
                     values = c("0" = "#F8766D", "1" = "#00BFC4"),
                     labels = c("No Disease", "Disease"))

```


Purpose: Understanding data distribution, correlations, and relationships between features and target variable. This helps us understand which features might be important predictors.

Inference:

- Age shows some separation between groups, but not extremely distinct

- Cholesterol levels show similar distributions across groups

- Gender shows clear patterns - males have higher heart disease prevalence

- Chest pain type is strongly correlated with heart disease presence

- The correlation matrix reveals complex relationships between clinical features


# Step 3: Data Preprocessing and Train-Test Split
```{r}
# Check for missing values
cat("Missing Values Summary:\n")
missing_summary <- sapply(heart, function(x) sum(is.na(x)))
print(missing_summary)

# Convert appropriate variables to factors
heart_processed <- heart %>%
  mutate(
    sex = as.factor(sex),
    cp = as.factor(cp),
    fbs = as.factor(fbs),
    restecg = as.factor(restecg),
    exang = as.factor(exang),
    slope = as.factor(slope),
    ca = as.factor(ca),
    thal = as.factor(thal),
    target = as.factor(target)
  )

# Create training and testing sets
set.seed(123)
train_index <- createDataPartition(heart_processed$target, p = 0.8, list = FALSE)
train_data <- heart_processed[train_index, ]
test_data <- heart_processed[-train_index, ]

cat("Training set size:", nrow(train_data), "\n")
cat("Testing set size:", nrow(test_data), "\n")
cat("\nTraining set distribution:\n")
print(table(train_data$target))
cat("\nTesting set distribution:\n")
print(table(test_data$target))
```


Purpose: Prepare data for modeling by handling categorical variables and creating proper train-test splits to evaluate model performance on unseen data.

Inference:

- No missing values in the dataset

- Successfully converted categorical variables to factors

- Created 80-20 train-test split maintaining similar class distribution

- Training set: 242 observations, Testing set: 61 observations

- Both sets have balanced class distribution for reliable evaluation


# Step 4: Decision Tree Implementation
```{r}
# Train Decision Tree with cross-validation
set.seed(123)
dt_control <- trainControl(
  method = "cv",  # "repeatedcv", "boot"
  number = 10,
  # repeats = 5,
  classProbs = TRUE,
  summaryFunction = twoClassSummary
  # savePredictions = TRUE
)

# Convert target to factor with proper names for caret
train_data_caret <- train_data %>%
  mutate(target = ifelse(target == 1, "Disease", "NoDisease") %>% as.factor())
test_data_caret <- test_data %>%
  mutate(target = ifelse(target == 1, "Disease", "NoDisease") %>% as.factor())

# Train Decision Tree
dt_model <- train(
  target ~ .,
  data = train_data_caret,
  method = "rpart",  # "C5.0",  "ctree", "tree",
  trControl = dt_control,
  tuneLength = 10,
  metric = "ROC"
)

# Make predictions
dt_pred <- predict(dt_model, test_data_caret)
dt_prob <- predict(dt_model, test_data_caret, type = "prob")[, "Disease"]

# Evaluate performance
dt_cm <- confusionMatrix(dt_pred, test_data_caret$target, positive = "Disease")
dt_accuracy <- dt_cm$overall["Accuracy"]
dt_sensitivity <- dt_cm$byClass["Sensitivity"]
dt_specificity <- dt_cm$byClass["Specificity"]

cat("Decision Tree Performance:\n")
print(dt_cm)

# Plot the decision tree
rpart.plot(dt_model$finalModel, 
           main = "Decision Tree for Heart Disease Prediction",
           box.palette = "RdYlGn",
           shadow.col = "gray",
           nn = TRUE)

# Feature importance from decision tree
dt_importance <- varImp(dt_model)$importance
dt_importance$Feature <- rownames(dt_importance)
dt_importance <- dt_importance[order(-dt_importance$Overall), ]

ggplot(dt_importance, aes(x = reorder(Feature, Overall), y = Overall)) +
  geom_col(fill = "steelblue", alpha = 0.8) +
  coord_flip() +
  labs(title = "Decision Tree - Feature Importance",
       x = "Features", y = "Importance Score") +
  theme_minimal()
```


Purpose: Implement and evaluate a single Decision Tree model. This gives us a baseline understanding of how a simple, interpretable model performs on our data.

Inference:

- The decision tree identifies key splitting features for heart disease prediction

- Feature importance shows which variables the tree considers most predictive

- The tree structure provides complete interpretability - we can trace every decision

- Performance metrics give us a baseline for comparison with Random Forest


# Step 5: Random Forest Implementation
```{r}

# Train Random Forest with cross-validation
set.seed(123)
rf_control <- trainControl(
  method = "cv",
  number = 10,
  classProbs = TRUE,
  summaryFunction = twoClassSummary
)

# Train Random Forest
rf_model <- train(
  target ~ .,
  data = train_data_caret,
  method = "rf",
  trControl = rf_control,
  tuneLength = 5,
  metric = "ROC",
  ntree = 500,
  importance = TRUE
)

# Make predictions
rf_pred <- predict(rf_model, test_data_caret)
rf_prob <- predict(rf_model, test_data_caret, type = "prob")[, "Disease"]

# Evaluate performance
rf_cm <- confusionMatrix(rf_pred, test_data_caret$target, positive = "Disease")
rf_accuracy <- rf_cm$overall["Accuracy"]
rf_sensitivity <- rf_cm$byClass["Sensitivity"]
rf_specificity <- rf_cm$byClass["Specificity"]

cat("Random Forest Performance:\n")
print(rf_cm)

# Plot Random Forest error rate
rf_final <- rf_model$finalModel
plot(rf_final, main = "Random Forest Error Rate")
legend("topright", 
       legend = colnames(rf_final$err.rate), 
       col = 1:ncol(rf_final$err.rate), 
       lty = 1, 
       cex = 0.8)

# FIXED: More robust feature importance extraction
# Method 1: Direct from randomForest object
rf_importance <- as.data.frame(rf_final$importance)
rf_importance$Feature <- rownames(rf_importance)

# Use MeanDecreaseGini for importance (most common for classification)
if("MeanDecreaseGini" %in% colnames(rf_importance)) {
  rf_importance$Importance_Score <- rf_importance$MeanDecreaseGini
} else {
  # If MeanDecreaseGini doesn't exist, use the first numeric column
  numeric_cols <- sapply(rf_importance, is.numeric)
  if(any(numeric_cols)) {
    rf_importance$Importance_Score <- rf_importance[, which(numeric_cols)[1]]
  } else {
    stop("No numeric importance scores found")
  }
}

# Order by importance
rf_importance <- rf_importance[order(-rf_importance$Importance_Score), ]

# Create the plot
ggplot(rf_importance, aes(x = reorder(Feature, Importance_Score), y = Importance_Score)) +
  geom_col(fill = "darkgreen", alpha = 0.8) +
  coord_flip() +
  labs(title = "Random Forest - Feature Importance",
       x = "Features", y = "Importance Score") +
  theme_minimal()

# Print the importance scores
cat("\nRandom Forest Feature Importance:\n")
print(rf_importance[, c("Feature", "Importance_Score")])
```


Purpose: Implement Random Forest to see how an ensemble method improves upon the single Decision Tree through bagging and feature randomness.

Inference:

- Random Forest typically shows more stable feature importance rankings

- The error rate plot shows how the model converges as more trees are added

- We can observe if the ensemble approach provides better generalization

- The feature importance might differ from the single tree due to ensemble averaging


# Step 6: Comprehensive Model Comparison
```{r}
# Create performance comparison
performance_comparison <- data.frame(
  Model = c("Decision Tree", "Random Forest"),
  Accuracy = c(dt_accuracy, rf_accuracy),
  Sensitivity = c(dt_sensitivity, rf_sensitivity),
  Specificity = c(dt_specificity, rf_specificity),
  AUC = c(auc(roc(test_data_caret$target, dt_prob)), 
          auc(roc(test_data_caret$target, rf_prob)))
)

cat("Performance Comparison:\n")
performance_comparison %>% kable(digits = 4) %>% kable_styling()

# Visualization of performance metrics
comparison_long <- performance_comparison %>%
  pivot_longer(cols = -Model, names_to = "Metric", values_to = "Value")

ggplot(comparison_long, aes(x = Metric, y = Value, fill = Model)) +
  geom_bar(stat = "identity", position = "dodge", alpha = 0.8) +
  labs(title = "Model Performance Comparison",
       x = "Performance Metrics", y = "Score") +
  theme_minimal() +
  scale_fill_manual(values = c("Decision Tree" = "steelblue", 
                              "Random Forest" = "darkgreen"))

# ROC Curve comparison
dt_roc <- roc(test_data_caret$target, dt_prob)
rf_roc <- roc(test_data_caret$target, rf_prob)

plot(dt_roc, col = "blue", main = "ROC Curve Comparison")
lines(rf_roc, col = "green")
legend("bottomright", 
       legend = c(paste("Decision Tree (AUC =", round(auc(dt_roc), 3), ")"),
                 paste("Random Forest (AUC =", round(auc(rf_roc), 3), ")")),
       col = c("blue", "green"), lwd = 2)

# FIXED: Feature importance comparison
# First, ensure both importance dataframes have consistent structure
dt_importance_clean <- dt_importance %>%
  select(Feature, DT_Importance = Overall) %>%
  mutate(DT_Importance = as.numeric(DT_Importance))

rf_importance_clean <- rf_importance %>%
  select(Feature, RF_Importance = Importance_Score) %>%
  mutate(RF_Importance = as.numeric(RF_Importance))

# Merge the importance data
importance_comparison <- merge(dt_importance_clean, rf_importance_clean, 
                              by = "Feature", all = TRUE)

# Convert to long format for plotting
importance_long <- importance_comparison %>%
  pivot_longer(cols = c(DT_Importance, RF_Importance), 
               names_to = "Model", 
               values_to = "Importance") %>%
  mutate(Model = ifelse(Model == "DT_Importance", "Decision Tree", "Random Forest"))

# Plot feature importance comparison
ggplot(importance_long, aes(x = reorder(Feature, Importance), y = Importance, fill = Model)) +
  geom_bar(stat = "identity", position = "dodge", alpha = 0.8) +
  coord_flip() +
  labs(title = "Feature Importance Comparison",
       x = "Features", y = "Importance Score") +
  theme_minimal() +
  scale_fill_manual(values = c("Decision Tree" = "steelblue", 
                              "Random Forest" = "darkgreen"))

# Print top features from each model
cat("\nTop 5 Features - Decision Tree:\n")
print(head(dt_importance_clean[order(-dt_importance_clean$DT_Importance), ], 5))

cat("\nTop 5 Features - Random Forest:\n")
print(head(rf_importance_clean[order(-rf_importance_clean$RF_Importance), ], 5))
```


Purpose: Directly compare both models across multiple metrics to understand their relative strengths and weaknesses on this specific dataset.

Inference:

- We can see which model performs better on accuracy, sensitivity, specificity

- AUC comparison shows which model has better overall discriminative power

- Feature importance comparison reveals if both models agree on important predictors

- The trade-off between interpretability (DT) and performance (RF) becomes clear



# Step 7: Statistical Significance Testing
```{r}
# Cross-validation for statistical comparison
set.seed(123)
cv_control <- trainControl(
  method = "repeatedcv",
  number = 10,
  repeats = 5,
  classProbs = TRUE,
  summaryFunction = twoClassSummary,
  savePredictions = TRUE
)

# Train both models with repeated CV
dt_cv <- train(
  target ~ .,
  data = train_data_caret,
  method = "rpart",
  trControl = cv_control,
  tuneLength = 5,
  metric = "ROC"
)

rf_cv <- train(
  target ~ .,
  data = train_data_caret,
  method = "rf",
  trControl = cv_control,
  tuneLength = 3,
  metric = "ROC",
  ntree = 500
)

# Compare model performance statistically
cv_comparison <- resamples(list(
  "Decision Tree" = dt_cv,
  "Random Forest" = rf_cv
))

cat("Cross-Validation Performance Summary:\n")
summary(cv_comparison)

# Visualization of CV results
bwplot(cv_comparison, main = "Cross-Validation Performance Comparison")

# Statistical significance test
cv_diff <- diff(cv_comparison)
cat("\nStatistical Significance of Differences:\n")
summary(cv_diff)
```

Purpose: Determine if the performance differences between models are statistically significant using rigorous cross-validation and hypothesis testing.

Inference:

- Repeated cross-validation provides more reliable performance estimates

- We can see if Random Forest's potential performance gains are statistically significant

- The results help us make data-driven decisions about model selection

- Understanding variance in performance across folds informs about model stability


# Step 8: Final Conclusions and Recommendations
```{r}
# Final model selection and insights
cat("=== FINAL COMPARISON AND RECOMMENDATIONS ===\n\n")

cat("1. PERFORMANCE SUMMARY:\n")
cat("   Decision Tree - Accuracy:", round(dt_accuracy, 3), 
    "AUC:", round(auc(dt_roc), 3), "\n")
cat("   Random Forest - Accuracy:", round(rf_accuracy, 3), 
    "AUC:", round(auc(rf_roc), 3), "\n\n")

cat("2. KEY DIFFERENCES OBSERVED:\n")
cat("   - Interpretability: Decision Tree wins (fully transparent)\n")
cat("   - Performance: Random Forest typically better (ensemble advantage)\n")
cat("   - Stability: Random Forest more robust to noise\n")
cat("   - Training time: Decision Tree faster\n")
cat("   - Feature importance: Both provide insights, RF more reliable\n\n")

cat("3. WHEN TO CHOOSE EACH MODEL:\n")
cat("   Choose Decision Tree when:\n")
cat("   * Interpretability is crucial (medical, legal domains)\n")
cat("   * Computational resources are limited\n")
cat("   * Need quick baseline model\n")
cat("   * Data relationships are relatively simple\n\n")

cat("   Choose Random Forest when:\n")
cat("   * Maximum accuracy is required\n")
cat("   * Dealing with complex feature interactions\n")
cat("   * Data has noise or missing patterns\n")
cat("   * Reliable feature importance is needed\n")
cat("   * Model stability is important\n\n")

cat("4. HEART DISEASE SPECIFIC INSIGHTS:\n")
# Top features from both models
top_features_dt <- head(dt_importance$Feature, 3)
top_features_rf <- head(rf_importance$Feature, 3)

cat("   Decision Tree top features:", paste(top_features_dt, collapse = ", "), "\n")
cat("   Random Forest top features:", paste(top_features_rf, collapse = ", "), "\n")
cat("   Common important features suggest robust clinical indicators\n")

# Final visualization: Model strengths
strengths_df <- data.frame(
  Aspect = rep(c("Interpretability", "Accuracy", "Speed", "Robustness"), 2),
  Score = c(9, 6, 9, 5, 3, 9, 4, 9),  # Scores 1-10
  Model = rep(c("Decision Tree", "Random Forest"), each = 4)
)

ggplot(strengths_df, aes(x = Aspect, y = Score, fill = Model)) +
  geom_bar(stat = "identity", position = "dodge", alpha = 0.8) +
  labs(title = "Model Strengths Comparison",
       x = "Performance Aspect", y = "Relative Score (1-10)") +
  theme_minimal() +
  scale_fill_manual(values = c("Decision Tree" = "steelblue", 
                              "Random Forest" = "darkgreen")) +
  ylim(0, 10)
```






