---
title: "Bike Sharing Prediction: Linear Regression vs Neural Network"
author: "Data Science Training"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    df_print: paged
    code_folding: show
  pdf_document:
    toc: true
    number_sections: true
---

# Bike Sharing Prediction: Linear Regression vs Neural Network

## Introduction

This analysis compares traditional Linear Regression with H2O Neural Network for predicting bike sharing demand using the UCI Bike Sharing Dataset.
The dataset contains **hourly** and **daily records of bike rentals** along with **environmental** and **seasonal covariates**.  

### **Key Variables**

- **instant:** Record index  
- **dteday:** Date of observation  
- **season:** Categorical variable representing the season (**winter**, **spring**, **summer**, **fall**)  
- **yr:** Year indicator (**0 = 2011**, **1 = 2012**)  
- **mnth:** Month of the year  
- **holiday:** Indicates whether the day is a **holiday**  
- **weekday:** Day of the week  
- **workingday:** Indicator for **non-holiday weekdays**  
- **weathersit:** Weather situation (**clear**, **mist**, **light rain/snow**)  
- **temp:** **Normalized temperature** in Celsius  
- **atemp:** **Normalized “feels-like” temperature**  
- **hum:** **Humidity** level  
- **windspeed:** **Normalized wind speed**  

### **Target and Related Variables**

- **casual:** Number of **casual users**  
- **registered:** Number of **registered users**  
- **cnt:** **Total bike rentals**, the **main target variable** for prediction 

These features combine **temporal**, **environmental**, and **behavioral** factors, making the dataset **ideal for regression tasks** such as **predicting the number of bike rentals** based on **weather** and **time-related covariates**.

## Step 1: Setup and Data Loading

```{r setup, message=FALSE, warning=FALSE}
# Load required libraries
library(h2o)
library(ggplot2)
library(dplyr)
library(tidyr)
library(caret)
library(GGally)
library(gridExtra)
library(knitr)
library(kableExtra)
library(patchwork)

# Load Bike Sharing Dataset from UCI
url <- "https://archive.ics.uci.edu/ml/machine-learning-databases/00275/Bike-Sharing-Dataset.zip"
temp <- tempfile()
download.file(url, temp)
unzip(temp, exdir = ".")
day_data <- read.csv("day.csv")

# Clean up
unlink(temp)
rm(temp)

cat("Bike Sharing Dataset loaded successfully\n")
cat("Dimensions:", dim(day_data), "\n")
cat("Variables:", names(day_data), "\n")

```
Purpose: Load the Bike Sharing dataset and initialize necessary libraries for the analysis.

Inference: Dataset successfully loaded with daily bike sharing records including weather, seasonal, and temporal features for demand prediction.

## Step 2: Data Exploration and Preprocessing
```{r }
# Display dataset structure
cat("Dataset Structure:\n")
str(day_data)

# Check for missing values
cat("\nMissing Values Summary:\n")
print(colSums(is.na(day_data)))

# Summary statistics
cat("\nSummary Statistics:\n")
summary(day_data)

# Target variable distribution
cat("\nTarget Variable (cnt) Distribution:\n")
summary(day_data$cnt)

# Select relevant features for regression
features_selected <- c('season', 'yr', 'mnth', 'holiday', 'weekday', 'workingday', 
                      'weathersit', 'temp', 'atemp', 'hum', 'windspeed')
target <- 'cnt'

bike_data <- day_data[, c(features_selected, target)]

# Convert categorical variables to factors
bike_data$season <- as.factor(bike_data$season)
bike_data$yr <- as.factor(bike_data$yr)
bike_data$mnth <- as.factor(bike_data$mnth)
bike_data$holiday <- as.factor(bike_data$holiday)
bike_data$weekday <- as.factor(bike_data$weekday)
bike_data$workingday <- as.factor(bike_data$workingday)
bike_data$weathersit <- as.factor(bike_data$weathersit)

cat("\nProcessed dataset dimensions:", dim(bike_data), "\n")

# Pairwise scatter matrix for numerical variables
numerical_vars <- c('temp', 'atemp', 'hum', 'windspeed', 'cnt')
ggpairs(bike_data[, numerical_vars],
        mapping = ggplot2::aes(alpha = 0.6),
        upper = list(continuous = wrap('cor', size = 3)),
        lower = list(continuous = wrap('smooth', alpha = 0.3, size = 0.4)),
        diag = list(continuous = wrap('densityDiag', alpha = 0.5))) +
  theme_minimal() +
  labs(title = "Pairwise Relationships in Bike Sharing Data")

# Target variable distribution
p_target <- ggplot(bike_data, aes(x = cnt)) +
  geom_histogram(fill = "steelblue", alpha = 0.7, bins = 30) +
  labs(title = "Bike Sharing Count Distribution",
       x = "Total Bike Count", y = "Frequency") +
  theme_minimal()

p_target

```
Purpose: Explore data distribution, relationships between variables, and understand the target variable for regression analysis.

Inference: The bike count shows a relatively normal distribution with some right skew. Strong correlations exist between temperature-related features and bike demand.

## Step 3: Data Preparation
```{r }
# Split data into train and test sets
set.seed(123)
train_index <- createDataPartition(bike_data$cnt, p = 0.8, list = FALSE)
train_data <- bike_data[train_index, ]
test_data <- bike_data[-train_index, ]

cat("Data split completed:\n")
cat("Training set:", nrow(train_data), "rows\n")
cat("Test set:", nrow(test_data), "rows\n")

# Check target variable distribution in splits
cat("\nTarget variable summary in training set:\n")
summary(train_data$cnt)
cat("\nTarget variable summary in test set:\n")
summary(test_data$cnt)

```
Purpose: Prepare data for modeling with proper train-test splits to ensure reliable model evaluation.

Inference: Data successfully split with similar target variable distribution in both training and test sets, ensuring fair model comparison.

## Step 4: Linear Regression Model
```{r }
# Train linear regression model
lm_model <- lm(cnt ~ ., data = train_data)

# Model summary
cat("Linear Regression Model Summary:\n")
print(summary(lm_model))

# Make predictions on test set
lm_predictions <- predict(lm_model, newdata = test_data)

# Calculate performance metrics
lm_rmse <- sqrt(mean((test_data$cnt - lm_predictions)^2))
lm_mae <- mean(abs(test_data$cnt - lm_predictions))
lm_r2 <- 1 - (sum((test_data$cnt - lm_predictions)^2) / 
               sum((test_data$cnt - mean(test_data$cnt))^2))

cat("\nLinear Regression Performance on Test Set:\n")
cat("RMSE:", round(lm_rmse, 2), "\n")
cat("MAE:", round(lm_mae, 2), "\n")
cat("R-squared:", round(lm_r2, 4), "\n")

# Create results dataframe for visualization
results_df <- data.frame(
  Actual = test_data$cnt,
  Linear_Regression = lm_predictions
)

# Plot actual vs predicted values
p_lm_fit <- ggplot(results_df, aes(x = Actual, y = Linear_Regression)) +
  geom_point(alpha = 0.6, color = "steelblue") +
  geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed") +
  labs(title = "Linear Regression: Actual vs Predicted",
       subtitle = paste("RMSE:", round(lm_rmse, 2), "R²:", round(lm_r2, 4)),
       x = "Actual Bike Count", y = "Predicted Bike Count") +
  theme_minimal()

# Residual plot
results_df$lm_residuals <- test_data$cnt - lm_predictions
p_lm_residuals <- ggplot(results_df, aes(x = Linear_Regression, y = lm_residuals)) +
  geom_point(alpha = 0.6, color = "darkorange") +
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
  labs(title = "Linear Regression: Residuals vs Predicted",
       x = "Predicted Bike Count", y = "Residuals") +
  theme_minimal()

grid.arrange(p_lm_fit, p_lm_residuals, ncol = 2)


```
Purpose: Implement and evaluate traditional Linear Regression as a baseline model for bike sharing prediction.

Inference: Linear Regression provides a reasonable baseline with interpretable coefficients, showing how different features affect bike sharing demand.

## Step 5: H2O Neural Network Setup
```{r }
# Initialize H2O cluster
h2o.init(nthreads = -1, max_mem_size = "2G")

# Convert data to H2O frames
train_h2o <- as.h2o(train_data)
test_h2o <- as.h2o(test_data)

# Define features and target for H2O
features_h2o <- setdiff(names(train_data), target)

cat("H2O setup completed:\n")
cat("Features:", features_h2o, "\n")
cat("Target:", target, "\n")

```
Purpose: Initialize H2O environment and prepare data for neural network training.

Inference: H2O cluster successfully initialized with proper data conversion for neural network modeling.

## Step 6: Simple Neural Network Model
```{r }
# Train a simple neural network
nn_model <- h2o.deeplearning(
  x = features_h2o,
  y = target,
  training_frame = train_h2o,
  model_id = "bike_nn_simple",
  hidden = c(10, 5),  # Very simple architecture
  epochs = 50,
  activation = "Rectifier",
  stopping_rounds = 5,
  stopping_tolerance = 0.01,
  stopping_metric = "RMSE",
  seed = 123
)

# Model summary
cat("Neural Network Model Summary:\n")
print(nn_model)

# Make predictions
nn_predictions <- h2o.predict(nn_model, test_h2o)
nn_pred_df <- as.data.frame(nn_predictions)

# Calculate performance metrics
nn_rmse <- sqrt(mean((test_data$cnt - nn_pred_df$predict)^2))
nn_mae <- mean(abs(test_data$cnt - nn_pred_df$predict))
nn_r2 <- 1 - (sum((test_data$cnt - nn_pred_df$predict)^2) / 
               sum((test_data$cnt - mean(test_data$cnt))^2))

cat("\nNeural Network Performance on Test Set:\n")
cat("RMSE:", round(nn_rmse, 2), "\n")
cat("MAE:", round(nn_mae, 2), "\n")
cat("R-squared:", round(nn_r2, 4), "\n")

# Add neural network predictions to results
results_df$Neural_Network <- nn_pred_df$predict

# Plot actual vs predicted for neural network
p_nn_fit <- ggplot(results_df, aes(x = Actual, y = Neural_Network)) +
  geom_point(alpha = 0.6, color = "darkgreen") +
  geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed") +
  labs(title = "Neural Network: Actual vs Predicted",
       subtitle = paste("RMSE:", round(nn_rmse, 2), "R²:", round(nn_r2, 4)),
       x = "Actual Bike Count", y = "Predicted Bike Count") +
  theme_minimal()

# Residual plot for neural network
results_df$nn_residuals <- test_data$cnt - nn_pred_df$predict
p_nn_residuals <- ggplot(results_df, aes(x = Neural_Network, y = nn_residuals)) +
  geom_point(alpha = 0.6, color = "purple") +
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
  labs(title = "Neural Network: Residuals vs Predicted",
       x = "Predicted Bike Count", y = "Residuals") +
  theme_minimal()

grid.arrange(p_nn_fit, p_nn_residuals, ncol = 2)

```
Purpose: Train and evaluate a simple neural network using H2O for comparison with linear regression.

Inference: The neural network captures non-linear patterns in the data, potentially offering improved predictions over linear regression.

## Step 7: Model Comparison

```{r }
# Create comprehensive comparison
performance_comparison <- data.frame(
  Model = c("Linear Regression", "Neural Network"),
  RMSE = c(lm_rmse, nn_rmse),
  MAE = c(lm_mae, nn_mae),
  R_squared = c(lm_r2, nn_r2)
)

cat("Model Performance Comparison:\n")
print(performance_comparison)

# Visualization: Performance metrics comparison
metrics_long <- performance_comparison %>%
  pivot_longer(cols = -Model, names_to = "Metric", values_to = "Value")

p_metrics <- ggplot(metrics_long, aes(x = Model, y = Value, fill = Model)) +
  geom_bar(stat = "identity", alpha = 0.8) +
  facet_wrap(~Metric, scales = "free_y") +
  labs(title = "Model Performance Metrics Comparison",
       y = "Metric Value") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Visualization: Side-by-side predictions
comparison_long <- results_df %>%
  select(Actual, Linear_Regression, Neural_Network) %>%
  pivot_longer(cols = -Actual, names_to = "Model", values_to = "Predicted")

p_side_by_side <- ggplot(comparison_long, aes(x = Actual, y = Predicted, color = Model)) +
  geom_point(alpha = 0.5) +
  geom_abline(intercept = 0, slope = 1, color = "black", linetype = "dashed") +
  labs(title = "Side-by-Side Prediction Comparison",
       x = "Actual Bike Count", y = "Predicted Bike Count") +
  theme_minimal() +
  scale_color_manual(values = c("Linear_Regression" = "#F8766D", 
                               "Neural_Network" = "#88E1E3"))

# Visualization: Residuals comparison
residuals_long <- results_df %>%
  select(Linear_Regression, Neural_Network, lm_residuals, nn_residuals) %>%
  pivot_longer(cols = c(lm_residuals, nn_residuals), 
               names_to = "Model", values_to = "Residuals") %>%
  mutate(Predicted = ifelse(Model == "lm_residuals", Linear_Regression, Neural_Network),
         Model = ifelse(Model == "lm_residuals", "Linear Regression", "Neural Network"))

p_residuals_compare <- ggplot(residuals_long, aes(x = Predicted, y = Residuals, color = Model)) +
  geom_point(alpha = 0.5) +
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
  labs(title = "Residuals Comparison",
       x = "Predicted Bike Count", y = "Residuals") +
  theme_minimal() +
  scale_color_manual(values = c("Linear Regression" = "#F8766D", 
                               "Neural Network" = "#88E1E3"))

# Display all comparison plots
# grid.arrange(p_metrics, p_side_by_side, p_residuals_compare, ncol = 1)

p_metrics / p_side_by_side / p_residuals_compare

```


Purpose: Comprehensive comparison of both models across multiple metrics and visualizations.

Inference: Direct comparison reveals the strengths and weaknesses of each approach, helping students understand when each method might be preferable.

## Step 8: Prediction Analysis and Insights
```{r }
# Analyze prediction errors
results_df$abs_error_lm <- abs(results_df$lm_residuals)
results_df$abs_error_nn <- abs(results_df$nn_residuals)

# Error distribution by actual value
p_error_dist <- ggplot(results_df) +
  geom_point(aes(x = Actual, y = abs_error_lm, color = "Linear Regression"), alpha = 0.6) +
  geom_point(aes(x = Actual, y = abs_error_nn, color = "Neural Network"), alpha = 0.6) +
  labs(title = "Absolute Error Distribution by Actual Value",
       x = "Actual Bike Count", y = "Absolute Error") +
  theme_minimal() +
  scale_color_manual(name = "Model", 
                     values = c("Linear Regression" = "#F8766D", 
                               "Neural Network" = "#88E1E3"))

# Best and worst predictions analysis
results_df$combined_rank <- rank(rowMeans(cbind(rank(results_df$abs_error_lm), 
                                               rank(results_df$abs_error_nn))))

best_predictions <- head(results_df[order(results_df$combined_rank), ], 10)
worst_predictions <- head(results_df[order(-results_df$combined_rank), ], 10)

cat("\nBest Predictions (Lowest Combined Error):\n")
print(best_predictions[, c("Actual", "Linear_Regression", "Neural_Network", 
                          "abs_error_lm", "abs_error_nn")])

cat("\nWorst Predictions (Highest Combined Error):\n")
print(worst_predictions[, c("Actual", "Linear_Regression", "Neural_Network", 
                           "abs_error_lm", "abs_error_nn")])

# Performance improvement analysis
improvement_df <- data.frame(
  Scenario = c("RMSE Improvement", "MAE Improvement", "R² Improvement"),
  Linear_Regression = c(lm_rmse, lm_mae, lm_r2),
  Neural_Network = c(nn_rmse, nn_mae, nn_r2),
  Improvement = c(
    round((lm_rmse - nn_rmse) / lm_rmse * 100, 2),
    round((lm_mae - nn_mae) / lm_mae * 100, 2),
    round((nn_r2 - lm_r2) / lm_r2 * 100, 2)
  )
)

cat("\nPerformance Improvement Analysis (%):\n")
print(improvement_df)

p_improvement <- ggplot(improvement_df, aes(x = Scenario, y = Improvement, fill = Scenario)) +
  geom_bar(stat = "identity", alpha = 0.8) +
  labs(title = "Neural Network Performance Improvement Over Linear Regression",
       y = "Improvement (%)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

p_error_dist / p_improvement

```

Purpose: Deep dive into prediction patterns and model performance characteristics.

Inference: Analysis reveals specific scenarios where each model excels or struggles, providing practical insights for model selection.

## Step 9: Final Conclusions and Recommendations
```{r }
cat("=== FINAL MODEL COMPARISON AND RECOMMENDATIONS ===\n\n")

cat("LINEAR REGRESSION RESULTS:\n")
cat("- RMSE:", round(lm_rmse, 2), "\n")
cat("- MAE:", round(lm_mae, 2), "\n")
cat("- R-squared:", round(lm_r2, 4), "\n")
cat("- Strengths: Highly interpretable, fast training, good baseline\n")
cat("- Weaknesses: Limited to linear relationships, may miss complex patterns\n\n")

cat("NEURAL NETWORK RESULTS:\n")
cat("- RMSE:", round(nn_rmse, 2), "\n")
cat("- MAE:", round(nn_mae, 2), "\n")
cat("- R-squared:", round(nn_r2, 4), "\n")
cat("- Strengths: Captures non-linear patterns, flexible architecture\n")
cat("- Weaknesses: Less interpretable, longer training time, hyperparameter sensitive\n\n")

cat("PERFORMANCE COMPARISON:\n")
if(nn_rmse < lm_rmse) {
  improvement <- round((lm_rmse - nn_rmse) / lm_rmse * 100, 1)
  cat("- Neural Network shows", improvement, "% lower RMSE\n")
} else {
  improvement <- round((nn_rmse - lm_rmse) / nn_rmse * 100, 1)
  cat("- Linear Regression shows", improvement, "% lower RMSE\n")
}

cat("- R-squared difference:", round(abs(nn_r2 - lm_r2), 4), "\n\n")

cat("RECOMMENDATIONS FOR STUDENTS:\n")
cat("1. Start with Linear Regression as a baseline for any regression problem\n")
cat("2. Use Neural Networks when you suspect complex non-linear relationships\n")
cat("3. Consider interpretability requirements when choosing between models\n")
cat("4. Always validate both simple and complex models on holdout data\n")
cat("5. Neural networks often work better with larger datasets and more features\n")

# Save models for future use
model_path_nn <- h2o.saveModel(nn_model, path = getwd(), force = TRUE)
cat("\nNeural Network model saved to:", model_path_nn, "\n")

# Shutdown H2O cluster
h2o.shutdown(prompt = FALSE)


```
